---
title: "MCMC and HMC for Rao genetic data"
author: "Roger Day"
date: "November 4, 2015"
output: html_document
---

Recall the set-up for Rao's genetics data:

Animals that are crosses of AB/ab x  AB/ab are classified in 4 categories: 
\[\begin{array}{*{20}{c}}
  Y&{Genotype}&{\Pr (category|\theta )}&{\Pr (category|\pi )} \\ 
  {125}&{AB}&{(3 - 2\theta  + {\theta ^2})/4}&{\frac{1}{2} + \frac{\pi }{4}} \\ 
  {18}&{Ab}&{(2\theta  - {\theta ^2})/4}&{\frac{1}{4} - \frac{\pi }{4}} \\ 
  {20}&{aB}&{(2\theta  - {\theta ^2})/4}&{\frac{1}{4} - \frac{\pi }{4}} \\ 
  {34}&{ab}&{(1 - 2\theta  + {\theta ^2})/4}&{\frac{\pi }{4}} 
\end{array}\]
The data are:
```{r }
y <- y.rao <- c(125,18,20,34)  #Original data.

```

The likelihood function is:
\[\begin{gathered}
  ODL \propto [Y|\pi ] \propto {\left( {\frac{1}{2} + \frac{\pi }{4}} \right)^{{Y_1}}}{\left( {\frac{1}{4} - \frac{\pi }{4}} \right)^{{Y_2} + {Y_3}}}{\left( {\frac{\pi }{4}} \right)^{{Y_4}}}  \\
  \quad \quad \quad \quad \quad \quad  \propto \sum\limits_{{X_1} = 0}^{{Y_1}} {\frac{{n!}}{{{X_1}!{X}!{Y_2}!{Y_3}!{Y_4}!}}{{\left( {\frac{1}{2}} \right)}^{{X_1}}}{{\left( {\frac{1}{4} - \frac{\pi }{4}} \right)}^{{Y_2} + {Y_3}}}{{\left( {\frac{\pi }{4}} \right)}^{{X} + {Y_4}}}}   \\
  \quad \quad \quad \quad = \sum\limits_{{X_1} = 0}^{{Y_1}} {\frac{{n!}}{{{X_1}!{X}!{Y_2}!{Y_3}!{Y_4}!}}CDL(\pi ;{X_1},{X},{Y_2},{Y_3},{Y_4})}   \\ 
\end{gathered} \]

The code for the likelihood function and its log:
```{r }
odl <- odl.rao <- function(y,p)
{
  (1/2+p/4)^y[1] * ((1-p)/4)^(y[2]+y[3]) * (p/4)^y[4] 
}

log.odl <- function(y,p) {
  log(odl(y,p))
}

log.cdl <- log.cdl.rao <- function(ExpX, y, p)
{
  if(length(p) > 1)
    return(sapply(p, log.cdl, ExpX=ExpX, y=y))
  log(dmultinom(
    x=c(y[1]-ExpX, ExpX, y[2], y[3], y[4]), 
    prob = c(1/2, p/4, (1-p)/4, (1-p)/4, p/4))
  )
}

```
Plots of the ODL and its log:
```{r}
p.delta <- 0.01
p.range <- seq(.01,.99,by=p.delta)
plot (p.range, odl(y,p.range), type="l",
      main="likelihood function")
plot (p.range, log.odl(y,p.range), type="l",
      main="log(likelihood)")

```


## Gibbs sampling for this problem.

```{r}
gibbsIter = function(current) {
  x2Star = current["x2"]
  phiStar = current["phi"]
  x2New = rbinom(n = 1, size = y.rao[1], prob = (phiStar/4) / (1/2+phiStar/4))
  phiNew = rbeta(n = 1, shape1 = x2New + y.rao[4] + 1, shape2 = y.rao[2] + y.rao[3] + 1)
  return( c(x2=x2New, phi=phiNew))
}

phiInit = runif(1)
x2Init = sample(0:125, 1)    ###   round(runif(1, max = 125))

# gibbsIter(c(x2=x2Init, phi=phiInit))

chainLength = 100
chainData = matrix(NA, nrow=chainLength, ncol=2, 
                   dimnames = list(1:chainLength, c("x2", "phi")))
chainData[1, ] = c(x2=x2Init, phi=phiInit)
for( iter in 2:chainLength) {
  chainData[iter, ] = gibbsIter(chainData[iter - 1, ])
}

par(mfrow=c(2,1))
plot(1:chainLength, chainData[ , "x2"])
plot(1:chainLength, chainData[ , "phi"])
hist(chainData[ , "x2"])
plot(density(chainData[ , "phi"]))
```

Using *rstan* to do MCMC on this problem is problematic, because STAN does not handle unknowns with discrete support. Therefore STAN cannot handle *x2* directly. Here is an attempt (among many), but it will not compile.
```{r}
stanCodeForRaoExample = '
data {
  int<lower=0> J; // number of categories (4) 
  real yRao[J]; // mouse counts
  int X2;
}
parameters {
  real<lower=0> phi;
  real<lower=0> pVec[5];
}
transformed parameters {
  //Jp1 <- J + 1;
  X[1] <- yRao[1] - X2;
  X[2] <- X2;
  for (j in 3:(J+1))
    X[j] <- yRao[j-1];
}
generated quantities {
  int X[J+1];
}
model {
  pVec[1] <- 1.0/2;
  pVec[2] <- phi/4;
  pVec[3] <- 1/4-phi/4;
  pVec[4] <- 1/4-phi/4;
  pVec[5] <- phi/4;
  X ~ binomial(125, pVec);
}
'
# fit.rao <- stan(model_code = stanCodeForRaoExample, 
#                 iter = 10, chains = 1)

```
A similar problem arises with mixture models with a finite number of components. A solution for mixture models is given in stan-reference-2.5.0.pdf, Chapter 9. It works directly with the marginalized model, summing over the components. Here we try that approach. It will be quite inefficient.

```{r}
stanCodeForRaoExample = '
data {
  int<lower=1> N; // number of categories
  int y[N]; // observations
}
parameters {
  real<lower=0, upper=1> phi; // 
}
transformed parameters {
  simplex[N] theta; // 
  theta[1] <- 1.0/2;
  theta[2] <- phi/4;
  theta[3] <- 1/4-phi/4;
  theta[4] <- 1/4-phi/4;
  theta[5] <- phi/4;
}
model {
  real ps[K]; // values for each
  for (k in 1:y[1]) {
    ps[k] <- ;
  }
  increment_log_prob(log_sum_exp(ps));
  for (j in 2:4) {
  }
}


